Cache Oblivious B-trees

### 2011-12-09 paper.tex
adding my more completed rough draft

\newcommand{\lhyperceil}{\lceil\lceil}
\newcommand{\rhyperceil}{\rceil\rceil}
\newcommand{\lhyperfloor}{\lfloor\lfloor}
\newcommand{\rhyperfloor}{\rfloor\rfloor}

\newcommand{\Search}{\textsc{Search(k)}}
\newcommand{\Insert}{\textsc{Insert(k, v)}}
\newcommand{\Scan}{\textsc{Scan(a, b)}}
\newcommand{\Delete}{\textsc{Delete(k)}}
\newcommand{\Succ}{\textsc{Succ(k)}}
\newcommand{\Pred}{\textsc{Pred(k)}}
\newcommand{\Member}{\textsc{Member(k)}}
%\newcommand{\Range}{\textsc{Range-Query(k$_{i}$,k$_{j}$)}}

@@

This paper present a survey on cache-oblivious B-trees (CO B-trees). The CO B-tree
was first introduce by Bender et al in 2000. It is an external memory data structure meaning
that it optimizes the number of memory transfers if performs rather than the number of
CPU instructions that it executions. Unlike other external memory data structure, like
the widely use B-tree, the CO B-tree does not require any information about the
number of level in the machine's memory hierarchy, the block size and capacity at each level, or the 
relative speeds for memory transfer. Hence, the CO B-tree is ``cache-oblivious''.
Despite this, the CO B-tree is able to achieve the same
optimal search bound as the B-tree and near optimal insertion and deletion bounds as well.
In this paper we first describe the original CO B-tree and explain how it is able to achieve 
its memory performance. We then present and analyze three followup CO B-tree data structures
 that improved upon shortcoming of the original: the cache-oblivious string B-tree, which introduces support 
for variable length keys, the cache-oblivious streaming B-tree, which incorporates improvements to the
insertion bound at the expense of the search bound, and lastly, the cache-oblivious concurrent B-tree, which 
introduces support for concurrency.

@@

Most algorithms assume an idealized model of computation that consists of a
single CPU, which performs instructions, and memory, which is a flat array
that can be read and written. However, in reality memory is not a flat array
but rather a hierarchy from super fast on chip caches, to slower main memory,
and even all the way down to the disk. When a program requests data, the data
is found somewhere in the memory hierarchy and typically carried up the
hierarchy to the fastest memory possible layer when the CPU is able to access
it. Unfortunately, this fast memory often has small capacity and once it is
filled and new data is requests some old entries must be evicted. Bring data
from lower to higher levels in the memory hierarchy often takes a long time.
In fact, access times between the different layers of the hierarchy can differ
by many orders of magnitude and often dominate the overall runtime performance
of an algorithm. External memory algorithms are wary of this fact and aim to
solve problem by optimizing the number of memory transfers it performs rather
than the number of instructions it executes. They often try to exploit the
fact the memory hierarchies transfer data in blocks rather than individual
atomic units. Thus, external memory algorithms are designed to not only limit
the number of memory transfers but also preserve the locality of data so that
when a memory access is performed, its cost is amortized.

@@

Since External memory algorithms are primarily concerned with efficient access
to data, it is not surprise that most problems can be reduced to finding a
suitable data structure that efficiently implement certain operations. A well
known external memory data structure is the B-tree. At a high level, B-tree
work by making their nodes the same size as a memory block in the hierarchy.
Furthermore, the fanout in the tree (the number of children) is also
proportional to the block size as well. The B-tree incurs $O(log(N))$ memory
transfers for \Search{}, \Insert{}, and \Delete{} and is widely used in
databases and filesystems.

\subsection{DAM vs CO}

The B-tree has two theoretical shortcomings. First, the B-tree is analyzed in
a two-level memory model called the Disk Access Machine (DAM) \cite{Aggarwal}. However,
in reality computing architectures typically have memory hierarchies that are
much larger than two. The argument for the DAM model is that in practice only
one level in the memory hierarchy tends to dominate the performance and thus a
two-level memory model is sufficient for all intents and purposes. However,
this does mean that, theoretically, B-trees can begin to incur an increasing
number of memory transfers if the dominating level in the hierarchy changes.

Additionally, the B-tree expects to know the the block size in the memory
hierarchy in advanced. This means that the parameterized in code need to be
changed for every level in the memory hierarchy which can be very difficult
and error prone. Additionally, tuning this parameter may even be impossible in
a heterogeneous computing environment.

The shortcomings of the DAM model gave rise to the cache-oblivious (CO) model
of computation. In the CO model the algorithm / data structure has no
information about the memory hierarchy. It does not know anything about the
speed of any level or the block size. At a conceptual level, the CO model
allow algorithm designers to reason about the simple two-level model like DAM
but prove results for any arbitrary memory hierarchy. In short, algorithms
that perform well in a CO model should, in theory, perform well regardless of
the block size in a memory hierarchy and thus should gracefully scale to
multi-level hierarchies as well.

@@

In \cite{BenderDemainColton} Bender et al. introduce the cache-oblivious B-tree (CO B-tree). The
CO B-tree is similar to normal the B-tree except that it does not have the
block size of levels in the memory hierarchy as explicitly parameters.

The CO B-tree is able to achieve the optimal search bound of $\Theta(log(N))$
memory transfers, the optimal scan bound of $\Theta(N/B)$ memory transfers and
efficient insertion and deletion operations. The CO B-tree was a seminal data
structure that gave rise to many others in the future.

This paper provides a survey of these different data structures. We start by
analyzing the original CO B-tree and explain how it works. This includes a
detailed description of the packed memory array (PMA) and strongly
weight-balance binary search trees (SWSBT) that are integral to the CO
B-tree's performance. Following this discussion we provide an analysis of the
CO String B-tree,a CO B-tree with variable length keys, the CO Streaming
B-tree, a CO B-tree that improves the cost of insertion at the expense of
search, and the CO Concurrent B-tree (which allows multiple processes to
access the data at the same time).

@@

High level idea:

We want to have a B-tree that is cache oblivious. Okay, well, first we still
need a way to actually store our data. We should use a tree. But what type of
tree? Obviously it needs to be balanced but anything else? Yes, it needs to
maintain the property that the number of descendants are $O(d^h)$. Why? It has
to do with how we will ultimately layout the tree in memory, but we will get
to that later.

Okay, so, we find the tree that we want to use. Now what? Well, first of all,
we need to make sure that it supports all of the operations that we want while
maintain this key property. Lets start with searching. We will layout the tree
in a vEB layout. This is helpful for two reasons. First, it is a nice way to
store it in memory and second because it gives us a search bound for free. Now
we need to make sure that it supports all of the other operations while
maintain the key property. I can show you how it handles insertions and
delete. Bam!

Okay, so now, we have a data structure that holds the amount of memory that we
want and it has the property that we will use later. Great! Now what? Well, we
need to actually store this in memory some how. We said that we will use a vEB
layout but that is only half of the story. We want a data structure that
maintain the locality of as many elements as once while also keeping extra
space to allow for future inserts. This is seemingly contradictory but we can
do it!

How? we use a PMA. The main problem is what inserts and delete may cause
splits and merges which may cause huge parts and nodes of the tree to move
around. So we need to first show how it is able to handle splits and merges
efficiently in a cache oblivious manner. bam! done! Great!

Now what? We also have the problem of updating pointers. If the points are
close to one anther that is not a big deal. However, it is possible that the
nodes are really far apart. Close means they are in the same block and thus
only incur a single memory transfer. Far apart means they are more. To solve
this problem we use buffer nodes and that works in a weird way. However, this
only gives us a bad bound, can we do better? ****

Yes! We use indirection. Indirection allows us to use another array and this
makes out pointer updates much more efficient.

We can now perform inserts and deletions faster. Additionally, our scans are
super fast as well. Search have not changed at all because we still have our
vEB structure and thus we are okay. But, we can also do a little better. If we
are okay without having scans, we can improve inserts and deletion since we no
longer have to maintain our indirection array.

The important part is that our analysis above we never treat the block size as
a constant. It is always a variable in our analysis but we are able to achieve
the desired bounds. Thus, we have a cache-oblivious B-tree that does
everything that we want.

@@

\subsection{Motivation}
In order for B-trees to remain versatile, they must be able to accommodate various forms of data. However, traditional B-trees and string B-trees have suboptimal performance when variable length strings are used as keys. In \cite{BenderFaKu06}, Bender et al. describe 4 major issues:
\begin{itemize}
\item B-trees handle variable length keys poorly
\item string B-trees do not support compression, and the front compression used in many other B-trees is handled sub-optimally
\item range queries are handled sub-optimally by regular B-trees, since they result in random disk seeks
\item traditional and string B-trees assume fixed block-transfer size B at all levels of the memory hierarchy, resulting in sub-optimal performance over hierarchies with varied block-transfer sizes.
\end{itemize}
\subsection{Description}
They address these issues by designing a Cache Oblivious String B-tree.

@@

For COSB-trees, the centroid tree is built using $O(N/logN)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to the nearest power of a power of 2. This ensures trees of equal size on all \emph{levels of detail}, as the centroid tree is traversed.

@@

The goal of modifying front compression is to maintain some of the space benefits to compression while reducing the high decoding cost and keeping the size of D small. Bender et al.  provide a modification that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\ll{D}\gg)$ after compression. The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters, add $\pi_{i}$,$\sigma_{i}$ like before. Otherwise, add 0,$k_{i}$, which means add $k_{i}$ to the stream in its uncompressed form. By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.%todo: add something about why this isO(<<D>>)?
%todo: add another paragraph or two about the second iteration
\subsection{Operators}
The operations supported are

>>>>

\newcommand{\lhyperceil}{\lceil\lceil}
\newcommand{\rhyperceil}{\rceil\rceil}
\newcommand{\lhyperfloor}{\lfloor\lfloor}
\newcommand{\rhyperfloor}{\rfloor\rfloor}
\newcommand{\llangle}{\langle\langle}
\newcommand{\rrangle}{\rangle\rangle}

\newcommand{\Search}{\textsc{Search(k)}}
\newcommand{\Insert}{\textsc{Insert(k, v)}}
\newcommand{\Insertkonly}{\textsc{Insert(k)}}
\newcommand{\Scan}{\textsc{Scan(a, b)}}
\newcommand{\Delete}{\textsc{Delete(k)}}
\newcommand{\Succ}{\textsc{Succ(k)}}
\newcommand{\Pred}{\textsc{Pred(k)}}
\newcommand{\Member}{\textsc{Member(k)}}
\newcommand{\Range}{\textsc{Range-Query(k$_{i}$,k$_{j}$)}}

@@

This paper presents a survey of cache-oblivious B-tree data structures. Memory hierarchies . 
Since memory transfers often dominate the runtimes of an algorithm it makes sense to
develop data structure that are optimized not in terms of the number of instructions it executes; but rather, 
by the number of memory accesses and transfers it has to perform. Among these so called ``external-memory'' data structures,
the B-tree is widely used as it is able to perform \Search in an optimal number of memory transfers. 
However, a short coming of B-trees is that they require knowledge of the size of the memory blocks, thus 
it is a cache aware data structure. This gave rise to the notion of cache-oblivious algorithms, e.g. data structures
that perform well regardless of the size of the memory blocks or the number hierarchies. Bender et al. 
introduced the cache-oblivious B-tree which ...

Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

@@

@@

\subsection{CO vs CA}

@@

@@

@@

\subsection{Motivation}
In order for B-trees to remain versatile, they must be able to accommodate various forms of data. However, traditional B-trees and string B-trees have suboptimal performance when variable length strings are used as keys. In addition, string B-trees do not support compression, and commonly used front compression is handled sub-optimally. Regular B-trees and string B-trees are not cache oblivious, and thus incur poorer performance as described previously in *cite original b-tree paper*. Range queries have poor performance in B-trees. In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache Oblivious String B-tree (COSB-tree), which they argue works efficiently in these problem areas for variable length string keys. They also provide a modified front-compression scheme, which gives their approach the advantages of using less space and having improved locality within the compressed data.

Two experiments are performed, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation. The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches. The results of the experiment show that COSB-trees outperformed B-trees most of the time for random searches, range queries and random inserts. However, some point between 440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree, causing it to fall behind in this measurement. However, Bender et al. argue that the experiment was biased towards the B-tree implementation, since the trees are "young", and thus still well organized. The COSB-tree implementation is also compared to Berkeley DB **cite berkeley db**, and achieves performance comparable to Berkeley DB with tuned parameters. The results seem to show that there is a window in which large and small-block size B-trees perform poorly, and that COSB-trees may be a reasonable alternative to improve performance.
\subsection{Description}

@@

The centroid tree is built using $O(N/logN)$ keys, and the centroid tree is layed out in memory using a modified van Emde Boas (vEB) layout for weight-balanced binary trees. The \emph{hyperhyperfloors} of nodes are used as a guide for cutting the tree into equal-size subtrees for the vEB layout. The \emph{hyperhyperfloor} of x is denoted as $\lhyperfloor{}x\rhyperfloor{}$, and refers to rounding x down to the nearest power of a power of 2. This ensures trees of equal size on all \emph{levels of detail}, as the centroid tree is traversed.

The following approach is used to rebalance the centroid tree: Put all the elements in DFS or Euler tour order for the trie. Given this order, scan the trie to identify the centroid vertex, partition the trie into its respective up and down tries, and recurse on the up and down tries. The Euler tour order can be achieved via $O(\log{}N)$ scans, since the trie is only $O(\log{}N)$ deep. Thus updates have amortized cost of $O(1+\log^{2}N/B)$ memory transfers. Note that this doesn't include updating successor or predecessor pointers, which is described in more detail below.

The following approach is used for maintaining the vEB layout for centroid trees in the dynamic COSB-tree: first, weight balance is relaxed from a maximum weight difference of 2 between left and right subtrees, to a larger constant $c$. Thus a node $v$ only becomes out of balance every $\Omega(\textsc{weight(v)})$ insertions or deletions of descendants of v, reducing the frequency of rebuilds. The nodes are stored in a PMA structure, preserving order of the nodes using $\Theta(N)$space, apart from inserts and deletes. On insert or delete, the structure is scanned in both directions starting from the update point for sections of the array to rebalance based on a "density" threshold of the section. Rebalance here refers to evenly distributing the nodes across the section. To ensure that updating pointers isn't too expensive, any node included in the section must have its subtrees included. Specifically, the section is built up from the the lowest subtree level that contains or contained the update point, and neighboring subtrees are continuously added until either the density threshold is met or the entire current subtree layout at this "level of detail" is included. For example, suppose a node is inserted at the leaf of some lower recursive subtree $L_{i}$. To continue building up this section, $L_{i-1}$ or $L_{i+1}$ could be added. If all the lower recursive subtrees $L_{1}$,$L_{2}$,...,$L_{x}$ have been added, the top level tree $U$ is added. This corresponds to a subtree $L'_{j}$ one level above. Thus repeat this process until the density threshold is met, and then rebalance the tree. This approach depends on the fact that all the recursive subtrees at the same level of detail asymptotically have the same size, which Bender et al. prove about the hyperhyperfloor-based vEB layout described previously. This approach results in an amortized update cost of $O(1+(\log{}N)/B)$ memory transfers per update. In conjunction with the cost of rebuilds described above, this is only an additive $O(\log{}N/B)$ factor, so the overall bound is amortized $O(1+\log^{2}N/B)$ memory transfers.

Note that making the COSB-trees dynamic causes updates to successor and predecessor pointers in the centroid tree to be expensive. Thus the notion of ancestors is relaxed to the following (Lemma 10 in \cite{BenderFaKu06}): Let $\alpha$ be the parent node of $\gamma$ in the compressed trie. Then either $\alpha$ is a descendant of $\gamma$ or $\gamma$ is a descendant of $\alpha$. The proof is as follows: As you recursively traverse the trie, both $\alpha$ and $\gamma$ are in the up trie or down trie, until at some point one of them must be the centroid vertex. This is used to redefine what the predecessor/ancestor pointers refer to. The centroid may no loner point to the correct predecessor or successor with respect to the relevant subtrie, but one of the matched ancestors by construction must point to the correct predecessor/successor. Thus for the dynamic case, when dealing with successors and predecessors, $O(logN)$ ancestors' successor and predecessor pointers in addition to the successor and predecessor pointers of the node in question must be checked. However, this is an additive factor, so the same time bound for update described above is still achieved.

@@

The goal of modifying front compression is to maintain some of the space benefits to compression while reducing the high decoding cost and keeping the size of D small. Bender et al.  provide a modification for static COSB-trees that reduces the high decoding cost of front compression of D to $O(1+||k||B)$, with size $O(\llangle{D}\rrangle{})$ after compression. The modification works as follows: Let $c = 2 + \epsilon/2$, and let $k_{1}$ be the first key added to the compression stream. For each subsequent key $k_{i}$, if $k_{i}$ can be decoded with just the last $c||k_{i}||$ characters, add $\pi_{i}$,$\sigma_{i}$ like before. Otherwise, add 0,$k_{i}$, which means add $k_{i}$ to the stream in its uncompressed form. By construction, this scheme guarantees that any key $k_{j}$ can be decoded in $O(1+||k||B)$ memory transfers.%todo: add something about why this isO(<<D>>)?

However, the above modification falls short for anticipating insertions and deletions. The previous modification relies on each key being decoded within $c||k^{*}||$ elements of the compressed array, but this is no longer guaranteed when new keys can be inserted to the left of $k$. To make the locality-preserving front compression technique suitable for dynamic COSB-trees, Bender et al. modify the algorithm to include an extra copied prefix for each key. Consider the $3c||k||$ characters that would be to the left of a new key$k^{*}$ after insertion. Let $k'$ be the key at the end of this range. The idea is to prevent the insertion of $k^{*}$ from affecting any non-copied keys just to the right of $k'$, since this could potentially increase their decode cost. Compute the length $l$ of the largest common prefix (lcp) between $k^{*}$ and $k'$, and compare $l$ to all the other lcp's in this range. $l$ represents the minimum length for any lcp between $k^{*}$ and $k'$, since the keys are stored in lexicographic order. If there is a copied prefix of length at least $l$ or a copied key within this range, insert $k^{*}$ as usual. Inserting $k^{*}$ will have no negative effect in this respect, since the decode length is effectively reset by the copied key or prefix like before. However, if inserting $k^{*}$ causes the minimum length of the lcp's in this range to increase, we need to fix the compression. To do this, we update the lcp of the key directly to the right of $k^{*}$ and store an extra copy of this lcp with $k^{*}$ itself. In this way, we can charge the work of copying extra prefixes for a new key to the appropriate number of elements before it. This number is determined by the length of the extra copied prefix stored with the key.

\subsection{Operators}
%todo: figure out how to do range queries from this paper somehow...
The operations supported are dictated by the Dictionary Matching problem. These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{}, and \Range{}. The behavior of each operation for dynamic COSB-trees is described below. \Insertkonly{} and \Delete{} are also supported but not described here.

\Search{}: works as follows for a given key $k$. Compute the hash value for $k$ and call it $\rho$, and traverse the centroid tree of the COSB-tree using $\rho$ to obtain the successor $\rho_{s}$ and predecessor $\rho_{p}$ of $k$. Index into the hashdata layer and perform a sequential scan across the PMA structure from $\rho_{p}$  to $\rho_{s}$ to find $\rho$. On a match with value $\rho'$, index into the keydata layer and decode $\rho'$ to obtain the resulting key $k'$. If $k' = k$, return true. If $k' \ne k$ or $k$ was not found during the sequential scan return false.

\Pred{},\Succ{}: works in the same way as \Search{} until the sequential scan in the hashdata layer. When scanning the hashdata layer, look for $\rho_{p}$ or $\rho_{s}$, respectively, in place of $\rho$, and compare the appropriate keys in the keydata layer.

\Range{}: The paper doesn't say, but presumably this is very similar to \Search{}. Let $k_{b}$ be the beginning of the range and let $k_{e}$ be the end of the range. Hash the range keys and identify the predecessor of $\rho_{b}$ and the successor of $\rho_{e}$ using the centroid tree. Scan the hashdata layer across from  $\rho_{b,p}$ to  $\rho_{e,s}$. Only start noting the beginning of range when the first key with hash value after $\rho_{b}$, and stop once we reach the first key with hash value after $\rho_{e}$. Decode and return the set of keys within the given range in the keydata layer.

\subsection{Complexity}
\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability (w.h.p.). Traversing the centroid tree requires  $O(1+\log_{B}N)$ memory transfers, since the tree is built using$O(N/logN)$ keys and is balanced. Once in the hashdata layer, we are already within $O(logN)$ keys of $k$. Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers. Obtaining $k$ from the keydata layer requires no more memory transfers than the scan in the hashdata layer. The \Pred{} and \Succ{} operations are almost identical in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively. These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p.. Similarly, range queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers, where in this case $k'$ is the other end of the range and $Q$ is the set of keys within the given range.

\Insertkonly{} and \Delete{} w.h.p. require $O(1+\log_{B}N+\log^{2}N||k||/B)$ memory transfers. As mentioned previously, due to rebuilds updates have amortized cost of $O(1+(\log^{2}N)/B)$ memory transfers. The rest of the cost for insertions and deletions is due to having to find the correct key to delete or location for insert, which can be seen from the cost of \Search{}.
